Forecasting the old-age dependency ratio to determine the best pension age

### 2020-09-14 PensionAge.Rmd
Reorganized some material and rewrote some sections

## Population forecasting

@@

The Lee-Carter (LC) method has been widely used to forecast mortality rates across various countries, including Australia. The model proposed by @LC92 (but using our notation) is
\begin{equation}\label{eq:lc}
  \log m_{t}(x)=\mu(x)+\phi(x)\beta_{t}+\varepsilon_{t,x},
\end{equation}
where $m_{t}(x)$ is the central death rate (properly defined in Section \@ref(sec:data)) for age $x$ in year $t$. This model involves fitting a matrix of log central death rates with sets of age-specific constants, {$\mu(x)$} and {$\phi(x)$}, and a time-varying index $\beta_{t}$. $\phi(x)$ is a set of parameters that tell us the rate at which each age responds to changes in $\beta_{t}$. $\beta_{t}$ measures the general level of the log death rates; $\mu(x)$ is the general pattern across the age of the log mortality rate and $\varepsilon_{t,x}$ is an error term with zero mean and constant variance, reflecting the randomness that can't be captured by the model.

@@

## Product-ratio method

@@

## Stochastic population forecasting

We follow the general procedure proposed in @HB08, except that we use the product-ratio method rather than the HU method for mortality and net-migration.

A stochastic functional data model was applied to the five components of population change: female mortality rates, male mortality rates, female net migration, male net migration and fertility rates to simulate these five components into the future.

We use the HU method applied to log-fertility, and the product-ratio method applied to log-mortality and net-migration. From these models, sample paths are simulated for age-specific fertility rates, age- and sex-specific mortality rates and age- and sex-specific net migration. Then future population values are simulated using the demographic growth-balance equations. The future deaths and births in this equation are assumed to follow a Poisson distribution, with parameters as a function of future mortality and fertility rates. We bootstrap the residuals of the net-migration model to simulate future net-migration numbers (see @HB08 for the details). Hence, we can simulate $h$-step-ahead age- and sex-specific population numbers based on simulated births, deaths and net-migrants.

This simulation was repeated for years $t=n+1,\dots,n+h$ to obtain a sample path of the age-specific population. Many simulated population paths were obtained in this way.

@@

# Data {#sec:data}

Historical central death rates and start-year and mid-year populations of residents in Australia by sex and age in single years for age group 0--99 and 100+ have been sourced from the @HMD for 1921--2018. The start-year population for 2019 is also used to estimate net migration. Data for age-specific annual fertility rates by single years of age for 15--49 over the period 1950--2018 was obtained from the @ABS12. Even though annual mortality rates are available from 1921 to 2018, only data collected after 1950 is used for two reasons. First, the period between 1921--1950 has different mortality patterns from later years because of wars and epidemics [@HB08]. Since the model for this study is valid under the assumption that extrapolative methods can be used, it seemed best to delete the less relevant data, which might influence estimations of parameters if we include them. Second, fertility rates are only available from 1950, so the mortality rate was cut to fit the same period.

## Forecasting OADR {#sec:forecasting}

Even though the OADR is a simple function of the population age structure and pension age, the formula can't be directly applied to the simulated population array because the pension age changes over time. To overcome this problem, we assume that the pension age can only be adjusted at the beginning of a year for these calculations. Therefore, one pension age is applicable for a whole year and hence OADR is generated at the beginning of each year to show the trend. We will show the process of obtaining a one-step-ahead OADR (i.e., forecast of OADR in the year 2019), given a specific pension age. The same method can be applied to all years over the prediction horizon, given their corresponding pension age. Further, the fractional pension age causes practical issues in calculating OADR, since ages are recorded as integers. We will also show how this issue was handled in the example below.

Suppose the pension age for $n+1$ is $a.r$ where $a$ is the integer and $r$ is the remainder of pension age. Recall that for $h=1$, we can express the simulated population paths as a matrix with age in the row and simulation in the column as follows:
$$
  \bm{p}_{n+1}=[p_{n+1,m,i}] =
    \begin{bmatrix}
       p_{1,1} & \cdots\cdots & p_{1,1000} \\
      \vdots & \ddots & \vdots \\
      p_{100^+,1} & \cdots\cdots &p_{100^+,1000}
      \end{bmatrix}.
$$
When $r=0$, then the simulation of a population aged 15 to pension age is a subset of the above matrix from $15^{th}$ row to $a-1^{th}$ row, which is:
\begin{equation}
  \begin{bmatrix}
    p_{15,1} & \cdots\cdots & p_{15,1000} \\
    \vdots & \ddots & \vdots \\
    p_{a-1,1} & \cdots\cdots &p_{a-1,1000}
  \end{bmatrix}. \label{eq:population_mat}
\end{equation}
To calculate the total number of working-age, we take a column sum of the trimmed matrix \eqref{eq:population_mat}. The simulations of the total number of working-age become a vector:
\begin{equation}
  \bm{p}^{\text{work}}_{n+1} = \left[\sum_{m=15}^{a-1}p_{m,1},\sum_{m=15}^{a-1}p_{m,2},\dots,\sum_{m=15}^{a-1}p_{m,1000}\right] \label{eq:work}
\end{equation}
Using the same method, the simulations of the total number of aged pensioners in the population can be shown as:
\begin{equation}
  \bm{p}^{\text{aged}}_{n+1} = \left[\sum_{m=a}^{100^+}p_{m,1},\sum_{m=a}^{100^+}p_{m,2},\dots,\sum_{m=a}^{100^+}p_{m,1000}\right]\label{eq:aged}
\end{equation}

When $r\neq0$, the vector of the total number of working-age population in \eqref{eq:work} is added, and the vector of the total number of aged population in \eqref{eq:aged} is subtracted with an adjustment:
\begin{equation}
  \text{adjustment}=0.r\times100\%[p_{a,1},p_{a,2},\dots,p_{a,1000}]
\end{equation}
This adjustment is based on the assumption that the birthday of the population is uniformly distributed over the year. Then, $r\times100\%$ of the population with a recorded integer age of $a$ at beginning of the year is actually younger than $a.r$, so they are still under the pension age. In other words, $(1-r\times100\%)$ of the population with recorded integer age $a$ at the beginning of the year is older than $a.r$, so they should be counted as the aged population. The simulation vectors of total work age population and total aged population should then be:
\begin{align*}
  \bm{p}^{\text{work}}_{n+1}
    & = \left[p^{\text{work}}_{n+1,1},p^{\text{work}}_{n+1,2},\dots,p^{\text{work}}_{n+1,1000}\right]\\
    & = \left[\sum_{m=15}^{a-1}p_{m,1},\sum_{m=15}^{a-1}p_{m,2},\dots,\sum_{m=15}^{a-1}p_{m,1000}\right]+0.r\times100\%\left[p_{a,1},p_{a,2},\dots,p_{a,1000}\right]   \\
  \bm{p}^{\text{aged}}_{n+1}
    & = \left[p^{\text{aged}}_{n+1,1},p^{\text{aged}}_{n+1,2},\dots,p^{\text{aged}}_{n+1,1000}\right]\\
    & = \left[\sum_{m=a}^{100^+}p_{m,1},\sum_{m=a}^{100^+}p_{m,2},\dots,\sum_{m=a}^{100^+}p_{m,1000}\right]-0.r\times100\%\left[p_{a,1},p_{a,2},\dots,p_{a,1000}\right]
\end{align*}

After the adjustment, we can generate simulation vectors of the one-step-ahead OADR by dividing every element of $\bm{p}^{\text{work}}_{n+1}$ by the corresponding element of $\bm{p}^{\text{aged}}_{n+1}$, i.e.:
\begin{align*}
  \bm{\text{OADR}}_{n+1} & = \left[\text{OADR}_{n+1,1},\text{OADR}_{n+1,2},\dots,\text{OADR}_{n+1,1000}\right]\\
                  & = \left[\frac{p^{\text{work}}_{n+1,1}}{p^{\text{aged}}_{n+1,1}},\dots,\frac{p^{\text{work}}_{n+1,1000}}{p^{\text{aged}}_{n+1,1000}}\right]
\end{align*}
The average value of elements of $\bm{\text{OADR}}_{n+1}$ is the mean prediction of one-step-ahead OADR.

The same algorithm is applied to the simulation matrix of the population age structure to obtain $h$-step-ahead simulation vectors of OADR for $h=1,2,\dots,H$, where $H=50$ for this study (i.e., a prediction horizon of 2019 to 2068). Simulations of OADR over a 50 year prediction horizon by the matrix is summarized below, with the forecast year in rows and simulation in the columns:
$$
  \bm{\text{OADR}}^{\text{forecast}}=\left[\text{OADR}_{h,i}\right] =
  \begin{bmatrix}
    \text{OADR}_{n+1,1} & \cdots\cdots & \text{OADR}_{n+1,1000} \\
    \vdots & \ddots & \vdots \\
    \text{OADR}_{n+50,1} & \cdots\cdots & \text{OADR}_{n+50,1000}
  \end{bmatrix}.
$$

## Results

@@

We use the fitted model of fertility rates as an example to interpret the results of fitted models. It should be noted that models fitted to product and ratio series are less likely to be interpreted in an economic sense. Figure \@ref(fig:componentsfertility) shows the estimated mean age pattern $\widehat{\mu}(x)$, the first two estimated basis functions, $\widehat{\phi}_1(x)$ and $\widehat{\phi}_2(x)$, and first two estimated coefficients (i.e., estimated principal component scores $\widehat{\beta}_{t,1}$ and $\widehat{\beta}_{t,2}$) for fertility rates. The first term ($\widehat{\phi}_1(x)\times \widehat{\beta}_{t,1}$) accounts for the most variation in fertility, so a relatively sensible interpretation can be given to the first term. In the plot of the first coefficient, estimated historical first principal component scores $\widehat{\beta}_{t,1}$ are shown in black for 1950--2018. It captures the general trend of fertility change over time. It indicates a sharp increase between 1950--1960, followed by a rapid decline in fertility rates from 1961. This increase of fertility is linked to the Post-World War II baby boom^[Between the years 1946 and 1964.] in Australia, resulting from the recovery from war and the Great Depression^[The Great Depression was a severe economic depression in the decade before World War II.]. The dramatic decrease from 1960 was due to the availability of contraception pills in Australia. A noticeable further reduction of fertility rates in the late 1960s may be due to a change in the law allowing abortion in some states in Australia.

@@

The first estimated basis function $\widehat{\phi}_1(x)$ measures the different effects of the trend over time across ages. It indicates the general pattern of fertility has little effect on females of age 15. This makes intuitive sense because most early pregnancies are accidental. Therefore, any of the previously discussed events has no effect on early pregnancies. In addition, there are two spikes in the first basic function, which suggests that the historical trend has a distinct impact on females aged around their 20s and 40s. Furthermore, the blue line is the fitted value of future coefficients (i.e., $\widehat{\beta}_{n+h,1}$, for $h=1,2,\dots,50$) and the yellow shading shows their 80% pointwise prediction interval according to the fitted ARIMA model of first coefficients $\left\{\widehat{\beta}_{t,1}\right\}$. Other terms further explain variation that has not been captured by the first term, but less economic explanations can be given for them.

>>>>

## Demographic growth-balance equations

## Old-Age Dependency Ratio

Using this notation, the OADR in year $t$ is given by
\begin{equation}\label{eq:oadr}
\text{OADR}_t = \frac{\displaystyle\sum_{x=15}^{\lfloor a_t\rfloor-1} P_t(x) + r_t P_t(\lfloor a_t\rfloor)}{\displaystyle\sum_{x\ge \lfloor a_t\rfloor} P_t(x) - r_t P_t(\lfloor a_t\rfloor)},
\end{equation}
where $a_t$ is the pension age and $r_t = a_t - \lfloor a_t \rfloor$ is the non-integer part of $a_t$. This assumes that the pension age can only be adjusted at the beginning of a year, and that the birthday of the population is uniformly distributed over the year.

## Forecasting demographic components

@@

The Lee-Carter (LC) method has been widely used to forecast mortality rates across various countries, including Australia. The model proposed by @LC92 (but using our notation) is
\begin{equation}\label{eq:lc}
  \log m_{t}(x)=\mu(x)+\phi(x)\beta_{t}+\varepsilon_{t,x},
\end{equation}
where $m_{t}(x)$ is the central death rate for age $x$ in year $t$. This model involves fitting a matrix of log central death rates with sets of age-specific constants, {$\mu(x)$} and {$\phi(x)$}, and a time-varying index $\beta_{t}$. $\phi(x)$ is a set of parameters that tell us the rate at which each age responds to changes in $\beta_{t}$. $\beta_{t}$ measures the general level of the log death rates; $\mu(x)$ is the general pattern across the age of the log mortality rate and $\varepsilon_{t,x}$ is an error term with zero mean and constant variance, reflecting the randomness that can't be captured by the model.

@@

### Product-ratio method {-}

@@

## Forecasting OADR

Following the general procedure proposed in @HB08, we model each of the five components of population change: female mortality rates, male mortality rates, female net migration, male net migration and fertility rates, and simulate these five components into the future.

We use the HU method applied to log-fertility, and the product-ratio method applied to log-mortality and net-migration (whereas @HB08 used the HU method for all five components).

From these models, sample paths are simulated for age-specific fertility rates, age- and sex-specific mortality rates and age- and sex-specific net migration. Then future population values are simulated using the demographic growth-balance equations. The future deaths and births in this equation are assumed to follow a Poisson distribution, with parameters as a function of future mortality and fertility rates. We bootstrap the residuals of the net-migration model to simulate future net-migration numbers (see @HB08 for the details). Hence, we can simulate $h$-step-ahead age- and sex-specific population numbers based on simulated births, deaths and net-migrants.

This simulation was repeated for years $t=n+1,\dots,n+h$ to obtain a sample path of the age-specific population. Many simulated population paths were obtained in this way. For each simulated population path, and each proposed set of pension age changes, we can simulate future OADR values using \eqref{eq:oadr}.

@@

# Results

## Data

Historical central death rates and start-year and mid-year populations of residents in Australia by sex and age in single years for age group 0--99 and 100+ have been sourced from the @HMD for 1921--2018. The start-year population for 2019 is also used to estimate net migration. Data for age-specific annual fertility rates by single years of age for 15--49 over the period 1950--2018 was obtained from the @ABSBirths. Even though annual mortality rates are available from 1921 to 2018, only data collected after 1950 is used for two reasons. First, the period between 1921--1950 has different mortality patterns from later years because of wars and epidemics [@HB08]. Since the model for this study is valid under the assumption that extrapolative methods can be used, it seemed best to delete the less relevant data, which might influence estimations of parameters if we include them. Second, fertility rates are only available from 1950, so the mortality rates were also truncated to the same period.

@@

## Component forecasts

We use the fitted model of fertility rates as an example to interpret the results of fitted models. It should be noted that models fitted to product and ratio series are less able to be interpreted in an economic sense. Figure \@ref(fig:componentsfertility) shows the estimated mean age pattern $\widehat{\mu}(x)$, the first two basis functions, $\widehat{\phi}_1(x)$ and $\widehat{\phi}_2(x)$, and the first two principal component scores, $\widehat{\beta}_{t,1}$ and $\widehat{\beta}_{t,2}$) for fertility rates. The first term ($\widehat{\phi}_1(x) \widehat{\beta}_{t,1}$) accounts for the most variation in fertility. The first principal component scores $\widehat{\beta}_{t,1}$, shown in black for 1950--2018, display a sharp increase between 1950--1960, followed by a rapid decline in fertility rates from 1961. This increase of fertility is linked to the Post-World War II baby boom in Australia, resulting from the recovery from war and the Great Depression. The dramatic decrease from 1960 was due to the availability of contraception pills in Australia. A noticeable further reduction of fertility rates in the late 1960s may be due to a change in the law allowing abortion in some states in Australia.

@@

The first basis function $\widehat{\phi}_1(x)$ measures the different effects of the time-trend across ages. It indicates the general pattern of fertility has little effect on females of age 15, most likely because such pregnancies are almost always accidental and usually prior to the mother taking contraception. The two "bumps" in the first basic function suggests that the historical trend has a distinct impact on females aged around their 20s and 40s. The second basis function largely concerns mothers aged over 30. The forecasts of future coefficients are shown as the blue lines in the lower panels with their 80% pointwise prediction intervals shown as blue shaded regions. These forecasts are obtained using automatically selected ARIMA models. In practice, we use six basis functions, but the four not shown have a much smaller effect on the resulting fertility forecasts and are much harder to interpret.

Similar models are used for the product and ratio of mortality rates, and sum and different of net migration rates.

## Population forecasts
